Phylotyper: {\it In silico} predictor of subtypes from gene sequences

### 2017-06-13 manuscript/response_to_reviewers.txt

Although the GitHub page is well laid out and easy enough to follow, please consider including some sort of example to follow along with. As a newcomer to the application it took me a little while to figure out how to get everything exactly right for an initial run. I think a very small example to follow and run that is located early in the GitHub instructions would really help to just get things going.

Response:
We agree with the reviewer's suggestion that a working example for new users will enhance the adoption and usability of Phylotyper. We have added an "Example" section to the README.md file (which is also the landing page for Phylotyper's GitHub repository: https://github.com/superphy/insilico-subtyping/#examples) along with example data that demonstrates how to 1) predict subtypes for schemes already packaged in Phylotyper and 2) how to add and run user-defined subtype schemes.

There lies a slight issue in the "user generated sub-type schemes" in that the results that you get with Phylotyper are highly dependant on the quality of the sub-type scheme provided. Since this tool will likely be used in outbreak situations and analyses, do you have any plans to address the quality of the sub-type scheme that was used? Do you plan on having a section in the GitHub page that allows users to send new sub-type schemes for validation and posting?

Response:
The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype predictor. To mitigate poor prediction from unsuitable or inaccurate inputs, we included in the build pipeline, checks to automatically test the performance of any submitted subtype scheme. These checks are now described in the Implementation section.

While these checks identify major breakdowns in the training set used by Phylotyper, we note there are no checks to identify missing data, specifically, missing subtypes or key gene sequences that define the range of the subtype sequence space. For any current general subtype predictor (i.e. not predictors that replicate PCR in silico, for instance), performance is dependent on the reference input set. In dealing with missing data, we demonstrate that Phylotyper is significantly more robust than a direct sequence similarity-based predictor. It is less likely to generate a Type-I error when encountering a subtype not found in the training set. The Phylotyper method estimates the uncertainty associated with a subtype prediction.

To enhance the users understanding of the checks that are provided in the build feature and the potential blindspots in creating their own subtype predictors with Phylotyper, we have greatly expanded the Implementation section and the README provided with the software. The added content describes the validations in Phylotyper, how they work and what they detect. In the Discussion section, we also provide guidelines for selecting sequences for an effective subtype predictor. We also included in the manuscript a new analysis that demonstrates how Phylotyper versus a BLAST-based predictor performs when an entire subtype class is removed from the training set and used as test input. This analysis and results have been added to the Results and Discussions sections.

The reviewer suggests having a means for users to submit subtypes for evaluation and posting to a shared repository. We agree that this would be a helpful system. We now mention in the manuscript and on the README, that users can contact us and submit new or modified subtype schemes. We will check the submitted schemes and add them to our Github repository for general use.

@@

----------------------------------------
Reviewer: 2

Comments to the Author
General comments:
Whiteside et al present here a novel python package for the prediction of bacterial molecular subtypes that are widely used in public health surveillance and outbreak analysis. The release of such a bioinformatic tools is highly valuable and timely with the democratization of WGS in microbiology laboratories worldwide. The article would benefit from some rewriting to provide more details to the reader regarding the use of Phylotyper and how to extend Phylotyper use to other bacteria than Escherichia coli.  Furthermore, a comparison of Phylotyper performance with other existing tools that the authors mention in the introduction would be welcome to assess the strength and weaknesses of this interesting phylogenetic approach.

Major comments:
Introduction:
1.      The authors should elaborate on the "current subtype systems" for readers that would not be familiar with clinical microbiology standards.

Response:
We added a high-level description of subtyping methods to the Introduction, which relates how subtyping can be categorized as phenotype- or DNA-based, with specific examples in each category. We feel this will give the readers an understanding of the breadth of subtyping methods. We also describe the characteristics of the subtypes that have been adapted for in silico prediction with WGS data. Our intention was to convey the wide range of subtyping approaches, and how certain subtyping approaches have been targeted for prediction with WGS data.

@@

Implementation:
2.      Phylotyper currently only has a scheme for Escherichia coli, which limits its use, and the interest resides in the possibility to build schemes for other bacteria. Therefore, it would be useful for the reader to have more information regarding how to build a new scheme in the contribution, and for example which existing resources of typing schemes could be used for main pathogens.

Response:
A detailed step-by-step account of how to run Phylotyper's automated build pipeline is provided in the Phylotyper package instructions. We reference and link to these instructions in the Implementation section. We also added to the Implementation section a description of the requirements in Phylotyper for new subtype schemes as well as the checks that the pipeline automatically performs to validate subtypes schemes created by users. In the Discussion section, we elaborate on the ideal sequence regions that should be targeted by users when developing a new WGS-based predictor of subtypes with Phylotyper.  To address the recommendation for providing existing typing schemes, we added in the Discussion section, some current examples of subtyping schemes in other pathogens outside of E. coli that can be predicted with Phylotyper. We did not systematically go through all main pathogens to identify the subtyping schemes that can be predicted with Phylotyper. As we state in our revised Discussion, the subtype schemes that can be linked biologically to a specific genome sequence region are the best candidates for WGS-based predictors. To compile a complete list of these type of subtypes requires domain knowledge in the pathogen (our lab works with E. coli, hence the initial focus in Phylotyper's resources on E. coli). We did identify and mention select examples in diverse pathogens to show that Phylotyper is not limited to E. coli. Phylotyper can work with any suitable subtype. Identifying a suitable subtype, however, requires knowledge of the subtype and its underlying biological mechanism. To help expand Phylotyper to other pathogens, we added a request in the manuscript and on the Phylotyper website for Microbiologists to send us either subtype schemes or additions/corrections to existing subtype schemes.  Subtype schemes that we receive will be packaged into the Phylotyper repository for download and general use.

@@

3.      Please provide more details regarding the “assumptions of the Phylotyper approach” and how/why they would be violated when building new subtype schemes.

Response: 
The Phylotyper assumptions are now outlined in the Implementation section. The revised Implementation section covers the basis of the assumptions, how they can be violated, and two built-in mechanisms we developed to alert users of possible violations.

Results:
4.      Whiteside et al compare here their tool with another BLAST based approach that they have developed for the comparison, and not methods published by other groups. The description of the method used for both the BLAST method and the cross validation is imprecise and would benefit from more being more detailed within the main article rather than in the supplementary material. Moreover, a comparison - at least theoretical - with other existing tools would strengthen the uniqueness of Phylotyper highlighted in the discussion.

Response:
Our main goal in the performance assessment was to compare how the prediction methods generalize to an independent dataset. There may exist differences in the reference training sets that underpin the tools and these differences may mask the true performance of the method. We wanted to evaluate the predictive capability of the methods themselves. SeroTypeFinder is provided as a web service. The software for SeroTypeFinder has not been made publicly available. Without access to the software, we cannot manipulate the training and test datasets to perform cross-validation analysis. The EcOH tool works directly on unassembled small reads and not assembled genome sequences. A direct comparison of the performance with these tools with Phylotyper is not possible. The intention of using a general BLAST-based approach is to demonstrate the benefit of Hidden State Prediction with a phylogenetic tree over approaches that directly match input sequences against a reference database. BLAST, as used in our assessment with a percent identity and coverage cutoff, is the same method used in the SeroTypeFinder tool. EcOH employs a more sophisticated algorithm for matching alleles in order to filter out spurious small reads that might align by chance. At its core, it is still matching alleles and does not develop a model of subtype allele evolution to define boundaries on the subtypes. We feel that our performance comparison to the generalized BLAST-based method does make a valid case for using HSP in subtype prediction. We did add a direct comparison of Phylotyper to V-typer for Stx genes. The V-typer comparison results are summarized in the Results section and provided fully in the Supplementary Information.

To address the comment that the BLAST and cross-validation method descriptions are imprecise, we have added more thorough descriptions of the BLAST method and the cross-validation analysis in the Results section. Our aim with the revised descriptions is to provide readers sufficient information to give them a complete understanding of the mechanics of the methods. Implementation details, such as percent identity cutoff or the number of iterations, remain in Supplementary Information. This was done to fit with the page constraints of the Application Note format.

As suggested, we also added a theoretical comparison of the tools to the Discussion; the theoretical comparison looks at use cases (assembled vs unassembled WGS), potential issues (false positive rates), and additional benefits outside of performance for the tools (prediction confidence values).

@@

Discussion:
5.      There is no data to support the claim that Phylotyper easily identifies novel subtype. Did the author try to remove all alleles of one subtype and compute the predictions to see if/how phylotyper creates a consistent subtype? Also, when a new subtype is identified, does Phylotyper includes it in his typing scheme automatically? What would be the recommended tests/further analysis recommended to researchers using Phylotyper in the case a new subtype is identified?

Response:
The point we were trying to convey was that more frequently, Phylotyper correctly handles novel subtypes than existing methods. In our empirical testing, we found Phylotyper makes fewer Type-I errors when encountering a novel allele of an existing subtype or an allele with an entirely new subtype not found in the reference training set. In the former case, Phylotyper returns non-significant/undetermined. We have clarified this point throughout the manuscript, removing references to Phylotyper identifying novel subtypes and replacing it with the more nuanced description that Phylotyper more frequently correctly classifies novel subtypes as a negative result or correctly classifies novel alleles with subtypes in the database. To strengthen this statement, we have as the reviewers suggest, added an analysis where we remove entire subtypes from the training set and calculated the false positive rate. Phylotyper was found to have a much lower false positive rate when classifying genes with unknown subtypes.

Phylotyper does not have the ability to create subtype schemes de novo or to add new subtype classes on the fly. We have not explored this potential application of the Phylotyper method. It would be an interesting avenue to explore as using a phylogenetic model of subtype evolution to define boundaries for new or existing subtypes might make a nice complement to existing de novo strategies like MLST that assign every new allele a separate subtype. However, our goal in this manuscript is to present an alternative genome-based subtype prediction method. De novo subtype creation is outside our original intended applications for Phylotyper.

@@

6.      “Subtypes are mainly used [..] the main uses of subtype information”. Please revise this paragraph, to allow a better logical flow. Also, Debroy et al have shown the discrepancy between between serotype-based typing and the genetic similarity of O antigen gene clusters, suggesting to replace serotyping by molecular typing. They did not particularly emphasize discrepancies between existing molecular typing scheme and whole genome data as suggested by Whiteside et al.

Response:

We have revised this paragraph and clarified our point on the suitability of Phylotyper for predicting subtypes from WGS; namely that Phylotyper helps identify discrepancies between a subtype and the genome regions that are used for prediction. We cite Debroy et al., as an example of how the associated genetic data and subtype schemes can have discrepancies.

@@

Figure 1.
6.      Please check the label that is cut on the side of the figure. Also, what does the letter at the end of each label correspond to - the expected subtype whereas the predicted subtype is color coded? Please better describe in the legend. Or remove if this is redundant information.

Response:
We removed the subtype letters at the end of the label; this was redundant information that was also conveyed by the color. Also we added a sentence in the figure description explaining the meaning of the colors. Labels are fixed so they fit in the window.

7.      The colors used by Phylotyper could be improved to increase readability of the output for the allele labels. There are several good packages to select colors within R or python.

Response:
We switched to the discriminatory palette "Set1" in R's RColorBrewer package. Phylotyper now uses RColorRamp and this palette to generate sufficient colors for all subtypes.

@@

Response:
The E. coli serotypes available in Phylotyper comprise two schemes: O- and H-type (prevously, incorrectly counted as one), resulting in five tested schemes. The reference to four has been removed.

----------------------------------------
Reviewer: 3

Comments to the Author
It's good to see a typing tool that is based on phylogenetic analysis. As public health practitioners still need to compare WGS data to old typing results, this tool will be useful to them.  Moreover, by putting typing into the framework of phylogenetic analysis, this tool can provide an interesting learning/visualization tool for epidemiologists and other public health practitioners to bridge the gap between subtypes (categorical data) and a tree (continuous data). The tool appears to be well implemented.
Some Specific Comments that the authors should address:
Major:
1) Please comment on what happens when an incomplete loci is found in the WGS data. Does the software warn the users of an partial alignment due to incomplete gene?

Response:
Complete or nearly complete loci sequences are needed for correctly placing the allele in the phylogenetic tree. To prevent incomplete sequences being used, we set coverage cutoffs of 90% in the BLAST detection step. No alleles with less than this coverage to one of the reference set sequences will pass the detection step. Users are notified when no gene copies are found. I have added a description of the requirement for complete gene sequences to the Implementation section.

@@

2) Please clarify how novel subtypes/alleles are identified and reported.

Response
Phylotyper is significantly more robust than a direct sequence similarity-based predictor in handling novel subtypes or alleles. We did not mean that Phylotyper performs unsupervised construction of new subtype schemes, rather, that Phylotyper has significantly lower false positive rates when performing prediction on an allele that is not in the reference set or an allele for which there is no example with the same subtype in the reference set (i.e. novel subtype). In the case of a novel subtype, Phylotyper would not assign the allele any subtype.  Compare this result with a sequence similarity approach, which, in most cases assign the novel subtype the closest subtype based on sequence.

To clarify this distinction, we have modified the text and removed all references of "Phylotyper identifying novel alleles", replacing it with the more specific statement.  We have also performed and included a subsequent analysis comparing the accuracy of the BLAST-based approach to the Phylotyper approach when the entire subtype group is withheld from the training set. This analysis examines the performance of the tools when encountering a novel subtype. It showed that Phylotyper is a marked improvement over the sequence-similarity approach, as Phylotyper more frequently returns a true negative result and not an incorrect subtype prediction.

@@

3) As there is a trend moving to core-genome or whole-genome MLST typing schemes, please comment if Phylotyper could be scaled up to handle that. How would horizontal gene transfer or recombination affect the Phylotyper results?

Response:
The Phylotyper approach is not designed for core-genome or whole-genome MLST schemes. The value that Phylotyper provides is that it can take input gene sequences and return predictions for subtypes that cannot be directly inferred from the gene sequences alone. For example, PCR or restriction digestion-based subtyping schemes are an abstraction of the underlying gene sequences and in most cases, cannot be directly inferred from an in silico genome sequence. We have demonstrated the ability of Phylotyper to accurately predict a PCR-based subtype (Stx1 and Stx2) as well as a phenotype-based scheme (O-type:H-type serotype).

Subtypes that are directly identified from the genome sequence do not require imputation from a software tool. In MLST, the combination of each unique allele sequence defines the type.  Potentially, and this is perhaps what the reviewer is referring to, a specific region of the genome that is smaller than the sequence regions used in an MLST, exists that can be used as a proxy to predict MLST subtype. The benefit, in this case, is that the entire genome would not need to be sequenced (or multiple alleles interrogated) in order to predict MLST type. There is a precedent for this type of approach.  The spa and coa genes in Staphylococcus aureus are used as gene markers to infer population variation.  The spa and coa types have noted concordance with MLST typing. While this is an interesting potential new application for Phylotyper, it is outside the scope of what we originally designed Phylotyper for.

Horizontal gene transfer or recombination could impact the result of Phylotyper in situations where the reference gene sequences used to predict subtype are only correlated with the subtype and not causative of the subtype. If the sequences used to predict the subtype scheme produce gene products that determine the subtype, such as the O-antigen processing genes for E. coli being used to predict O-group serotypes, the Phylotyper prediction should directly correspond to the subtype. In contrast, if sequences that are correlated but have no biological link are used, HGT could recombine a particular gene allele with a new subtype. We have made the readers aware of this potential issue and added to the Discussion section recommendations for users building subtype schemes to focus on genomic regions that have a direct biological role.

@@

Minor:
1) Typo in line 49: also not aslo

Response:
Change made.

2) Line 51 mentioned assumptions for building new subtype pipeline.  It'll be good to explain that or if space is an issue, point to the portion of the manual that explains these assumptions

Response:
We have added descriptions of the Phylotyper method assumptions to the Implementation section. The revised Implementation section covers the basis of the assumptions, how they can be violated, and two built-in mechanisms we developed to alert users of possible violations.

### 2017-06-13 manuscript/phylotyper_revised.tex
MODIFY: Chad's corrections added to document MODIFY: edits to response letter

WWhole-genome sequencing (WGS) is transforming the public health field by providing an efficient method for surveying bacterial populations.
The speed, discriminatory power and broad utility of WGS can improve surveillance and outbreak analysis.
Adoption of WGS in public health, however, requires transitioning of historical data with the new methods \citep{Jenkins2015}.
One of the workhorse methods in public health is subtyping.
Subtyping methods can broadly be categorized as phenotype-based or DNA-based.
Phenotype-based subtypes are, for example, interrogated by biochemical tests (biotyping), detection of surface antigens (serotyping) or susceptibility to bacteriophage (phagetyping) \citep{Wiedmann2002}.
Alternatively, DNA-based subtyping examines and classifies bacteria based on genetic content.
DNA-based subtypes use a variety of methods from PCR to Pulse Field Gel Electrophoresis to DNA sequencing to assign groups to bacteria \citep{Wiedmann2002}.
As a surveillance tool, subtypes provide a clear cut designation that is typically used to distinguish taxonomic groups and infer phenotypes.
A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain\citep{Jenkins2015}.
Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.
To predict from WGS data, these tools have targeted subtypes (either Phenotype-based or DNA-based systems) that can be predicted by sequence variation in a specific region or gene in the genome.
\textit{In silico} subtype prediction is not applied to subtyping methods that already directly use the DNA sequence to assign subtype such as Multi-Locus Sequence Typing (MLST). 
No inference or extrapolation is needed to integrate a direct DNA sequence-based method, like MLST, into a WGS system.
An example of a subtyping system that has been adapted for WGS is serotyping in \textit{Escherichia coli}.
The sequence of O-antigen processing genes in \textit{E. coli} is used to predict O-antigen group serotype in \citep{Joensen2015,Ingle2016}.
Another example is the Shiga toxin subtype (Stx); a DNA-based subtyping scheme generated using PCR.
The tool in \citep{CARRILLO2016} predicts Stx subtype by simulating PCR \textit{in silico}.

@@

Currently, the Phylotyper package includes the following subtype schemes for \emph{Escherichia coli}: Stx, intimin and serotype O- and H-types  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes (instructions are provided here: \url{https://github.com/superphy/insilico-subtyping}). 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
Checks are built-in to the \texttt{new} pipeline to validate the submitted reference set.
Each new subtype is subject to two tests and results are reported to the user.
The first test checks that the distribution of inter-patristic phylogenetic distances between instances of the same subtype is both smaller and distinct from subtypes that are different.
This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (see Supplementary Methods for details).
The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure S2).
These checks verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
There is no check that can confirm the training set covers all subtypes in a particular scheme; Phylotyper can only predict subtypes that are represented in the training set.
It is important that the user monitor schemes and update it as gaps are identified.
Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering an unknown subtype that has no representative in the training set.

@@

Phylotyper is a progression from the sequence-similarity approach that is the basis of current \textit{in silico} subtype prediction strategies.
To compare Phylotyper to a sequence similarity-based approach, we ran two validations that looked at how both methods perform when confronted with 1) gene sequence or 2) subtype class not present in the training set.
The first validation was a leave-one-out cross-validation test that iterated through each gene in the training set, retraining the prediction tools on a reduced training set that excludes the selected test gene, and then confirming if the retrained predictor could recover the subtype of the test gene.
This validation tests how the predictors perform when run on a distinct sequence that is not in the training set.
The second validation examined how the predictors perform when tested with a gene that has a subtype, not in the reference set.
In this validation, each subtype was iterated over and all genes that are assigned the subtype were removed from the training set.
In each iteration, we recorded the number of false-positive subtype assignments when the test sequences were used as input.
The correct response for the predictors was to return a negative result since the subtype does not exist in the training set.
For these assessments, we developed a sequence-similarity based tool that assigns putative subtypes using BLAST.
This generalized BLAST tool, based on the approach used in \citep{Joensen2015}, assigns a query sequence a subtype when the top BLAST match from an annotated reference database is above a pre-selected percent identity and alignment coverage cutoff.
Details how the assessment was conducted are available in Supplementary methods.
The assessment examined the five subtype schemes available in Phylotyper: Stx1, Stx2, Eae, H-type (FliC), O-type (Wzy \& Wzx).
When tasked with assigning a novel gene sequence not in the training set in the leave-one-out validation, Phylotyper consistently had higher precision than a top-BLAST-hit approach.
The average precision in Phylotyper was 0.99 versus 0.96 in the top-BLAST-hit approach (Supplementary Table S2).
The BLAST approach also often had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper.
Similarly, when entire subtype classes were withheld from the training set, Phylotyper had consistently lower false positive rates for all subtypes schemes tested; the average false positive rate in this test case was 0.11, while in the BLAST approach, the average false positive rate was 0.30.
A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set \citep{CARRILLO2016}.
The test Stx genes had sufficient surrounding DNA sequence to support \textit{in silico} PCR.
In total, 24 Stx gene sequences were tested with the V-typer tool and V-typer returned results for 7, all correct.
Phylotyper correctly predicted the subtype for all these genes.
Based on this level of recall, it appears conditions in the Stx subtype environment are challenging for simulated PCR.

@@

The main strategy currently in use by other \textit{in silico} tools for predicting subtypes is to use sequence similarity to annotated gene alleles.
Query genomes or genes are matched to alleles that are attributed the subtype phenotype or are correlated with the distribution of the subtype.
For example, SerotypeFinder uses BLAST to find the top matches based on sequence similarity to O-antigen processing genes for \textit{in silico} O-typing and the flagellin genes for H-typing \textit{E. coli} genes \citep{Jenkins2015}.  
O-type and H-type are transferred from the top matches to the queries provided they are above coverage and percent identity thresholds.
This general strategy of allele matching is also applied in the EcOH tool \citep{Ingle2016}, however, the EcOH tool can directly use unassembled sequence reads as input. The EcOH tool aligns reads to alleles linked to \textit{E. coli} O-types and H-types, and identifies the top candidates that have an alignment score above pre-defined thresholds.
Phylotyper is comparatively more robust as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set.
With the allele matching strategy, the reference set make-up can have a greater impact on performance.
When alleles or even subtypes are missing in the reference database, the sequence similarity approach more frequently generates false positive predictions.
Alternatively, Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make a Type-I error when encountering a novel subtype or allele. 
The empirical testing we performed demonstrated this behavior; 
the rate of false positive classifications was significantly lower than in a sequence-similarity approach in validations where we withheld an allele or an entire subtype from the training set and used it as a test input.
V-typer takes a distinct approach; it directly simulates the \textit{in vitro} wet-lab PCR procedure used to perform Stx subtyping \citep{CARRILLO2016}. 
V-typer's direct replication of a wet-lab method \textit{in silico} means it can only be applied to subtypes schemes that use PCR. 
Additionally, we found in our evaluation of Stx subtypes that it failed to generate predictions for most test cases.
From a methodology standpoint, Phylotyper has an additional benefit over current methods; the phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting results.
In comparison, there is no built-in mechanism in the sequence similarity approach to quantify the level of confidence in assigning alleles a subtype.

Subtypes are mainly used as a proxy for evolutionarily-related bacterial strain groups or to infer phenotypes.
A recent analysis of O-antigen serotypes and their associated O-antigen gene sequences in \textit{E. coli} found that the sequence data indicated several changes to the organization of the O-groups \citep{DebRoy2016}.
There are potentially other subtype schemes that would show discrepancies between genetic data and subtype grouping.
A tool that can evaluate the ability of a genotype to predict a subtype would be better equipped for developing the new subtype schemes or updating current schemes for WGS workflows.
Phylotyper's \texttt{add} pipeline tests subtype schemes for their predictive accuracy by returning an F-score based on a cross-validation assessment.
The validation verifies that the phylogeny generated by the training set sequences can be used to predict the subtypes with a high level of accuracy.
We showed in empirical tests that the farther a subtype is dissociated from the input gene's phylogeny, the lower the F-score computed in the Phylotyper \texttt{add} pipeline.
In addition to this subtype-level verification, Phylotyper also computes confidence scores for each individual prediction that reflect the rate of subtype change occurring at the input gene's phylogenetic locale.
If an input sequence falls in a region in the phylogenetic tree where the subtype is highly fluid due to evolution or poor subtype-genotype correlation, users would be notified in the confidence score and in the phylogenetic tree visualization output by Phylotyper. 
Phylotyper's ability to inform users about the level of agreement between the subtype assignments and genotype makes it uniquely capable of transitioning historical subtype data to new whole genome sequence-based systems.\vspace*{-10pt}

>>>>

WWhole-genome sequencing (WGS) is transforming the public health field by providing an efficient method for surveying bacterial populations.
The speed, discriminatory power and broad utility of WGS can improve surveillance and outbreak analysis.
Adoption of WGS in public health, however, requires transitioning of historical data with the new methods \citep{Jenkins2015}.
One of the workhorse methods in public health is subtyping.
Subtyping methods can broadly be categorized as phenotype-based or DNA-based.
Phenotype-based subtypes are, for example, interrogated by biochemical tests (biotyping), detection of surface antigens (serotyping) or susceptibility to bacteriophage (phagetyping) \citep{Wiedmann2002}.
Alternatively, DNA-based subtyping examines and classifies bacteria based on genetic content. DNA-based subtypes use a variety of methods from PCR and Pulse Field Gel Electrophoresis, to DNA sequencing to assign subtype designations to bacterial isolates \citep{Wiedmann2002}.
As a surveillance tool, subtypes provide a clear cut designation that is typically used to distinguish taxonomic groups and infer phenotypes.
A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain\citep{Jenkins2015}.
Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.
To predict from WGS data, these tools target subtypes (either Phenotype-based or DNA-based systems) that can be predicted by sequence variation in a specific region or gene in the genome.
\textit{In silico} subtype prediction is not applied to subtyping methods that already directly use the DNA sequence to assign subtype such as Multi-Locus Sequence Typing (MLST). 
No inference or extrapolation is needed to integrate a direct DNA sequence-based method, like MLST, into a WGS system.
An example of a subtyping system that has been adapted for WGS is serotyping of \textit{Escherichia coli}.
The sequence of O-antigen processing genes in \textit{E. coli} is used to predict O-antigen group serotype in \citep{Joensen2015,Ingle2016}.
Another example is the Shiga toxin subtype (Stx); a DNA-based subtyping scheme generated using PCR.
The tool in \citep{CARRILLO2016} predicts Stx subtype by simulating PCR \textit{in silico}.

@@

Currently, the Phylotyper package includes the following subtype schemes for \emph{Escherichia coli}: Stx, intimin and serotype O- and H-types  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes (instructions are provided here: \url{https://github.com/superphy/insilico-subtyping}). 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
Checks are built-in to the \texttt{new} pipeline to validate the submitted reference set.
Each new subtype is subject to two tests and results are reported to the user.
The first test checks that the distribution of inter-patristic phylogenetic distances between instances of the same subtype is both smaller and distinct from subtypes that are different.
This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (see Supplementary Methods for details).
The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure S2).
These checks verify that the phylogenetic grouping provided by the training set is predictive of the subtype.
There is no check that can confirm the training set covers all subtypes in a particular scheme; Phylotyper can only predict subtypes that are represented in the training set.
It is important that the user monitor schemes and update it as gaps are identified.
Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering an unknown subtype that has no representative in the training set.

@@

Phylotyper is a progression from the sequence-similarity approach that is the basis of current \textit{in silico} subtype prediction strategies.
To compare Phylotyper to a sequence similarity-based approach, we ran two validations that looked at how both methods perform when confronted with 1) gene sequence or 2) subtype class not present in the training set.
The first validation was a leave-one-out cross-validation test that iterated through each gene in the training set, retraining the prediction tools on a reduced training set that excludes the selected test gene, and then confirming if the retrained predictor could recover the subtype of the test gene.
This validation tests how the predictors perform when run on a distinct sequence that is not in the training set.
The second validation examined how the predictors perform when tested with a gene that has a subtype not in the reference set.
In this validation, each subtype was iterated over and all genes that were assigned the subtype were removed from the training set.
In each iteration, we recorded the number of false-positive subtype assignments when the test sequences were used as input.
The correct response for the predictors was to return a negative result since the subtype does not exist in the training set.
For these assessments, we developed a sequence-similarity based tool that assigns putative subtypes using BLAST.
This generalized BLAST tool, based on the approach used in \citep{Joensen2015}, assigns a query sequence a subtype when the top BLAST match from an annotated reference database is above a pre-selected percent identity and alignment coverage cutoff.
Details how the assessment was conducted are available in Supplementary Methods.
The assessment examined the five subtype schemes available in Phylotyper: Stx1, Stx2, Eae, H-type (FliC), O-type (Wzy \& Wzx).
When tasked with assigning a novel gene sequence not in the training set in the leave-one-out validation, Phylotyper consistently had higher precision than a top-BLAST-hit approach.
The average precision in Phylotyper was 0.99 versus 0.96 in the top-BLAST-hit approach (Supplementary Table S2).
The BLAST approach also had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper.
Similarly, when entire subtype classes were withheld from the training set, Phylotyper had consistently lower false positive rates for all subtypes schemes tested; the average false positive rate in this test case was 0.11, while in the BLAST approach, the average false positive rate was 0.30.
A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set \citep{CARRILLO2016}.
The test Stx genes had sufficient surrounding DNA sequence to support \textit{in silico} PCR.
In total, 24 Stx gene sequences were tested with the V-typer tool and V-typer returned results for 7, all correct.
Phylotyper correctly predicted the subtype for all these genes.
Based on this level of recall, it appears conditions in the Stx subtype environment are challenging for simulated PCR.

@@

The main strategy currently in use by other \textit{in silico} tools for predicting subtypes is to use sequence similarity to annotated gene alleles.
Query genomes or genes are matched to alleles that attribute the subtype, or that are correlated with the subtype.
For example, SerotypeFinder uses BLAST to find the top matches based on sequence similarity to O-antigen processing genes for \textit{in silico} O-typing and the flagellin genes for H-typing \textit{E. coli} genes \citep{Jenkins2015}.  
O-type and H-type are transferred from the top matches to the queries provided they are above coverage and percent identity thresholds.
This general strategy of allele matching is also applied in the EcOH tool \citep{Ingle2016}, however, the EcOH tool can directly use unassembled sequence reads as input. The EcOH tool aligns reads to alleles linked to \textit{E. coli} O-types and H-types, and identifies the top candidates that have an alignment score above pre-defined thresholds.
Phylotyper is comparatively more robust as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set.
With the allele matching strategy, the reference set make-up can have a greater impact on performance.
When alleles or even subtypes are missing in the reference database, the sequence similarity approach more frequently generates false positive predictions.
In contrast, Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make a Type-I error when encountering a novel subtype or allele. 
The empirical testing we performed demonstrated this behavior; 
the rate of false positive classifications was significantly lower than in a sequence-similarity approach in validations where we withheld an allele or an entire subtype from the training set and used it as a test input.
V-typer takes a distinct approach; it directly simulates the \textit{in vitro} wet-lab PCR procedure used to perform Stx subtyping \citep{CARRILLO2016}. 
V-typer's direct replication of a wet-lab method \textit{in silico} means it can only be applied to subtypes schemes that use PCR. 
Additionally, we found in our evaluation of Stx subtypes that it failed to generate predictions for most test cases.
From a methodology standpoint, Phylotyper has an additional benefit over current methods; the phylogenetic framework in Phylotyper provides a statistical likelihood for interpreting results.
In comparison, there is no built-in mechanism in the sequence similarity approach to quantify the level of confidence in assigning alleles a subtype.

Subtypes are mainly used as a proxy for evolutionarily-related bacterial strain groups or to infer phenotypes.
A recent analysis of O-antigen serotypes and their associated O-antigen gene sequences in \textit{E. coli} found that the sequence data indicated several changes to the organization of the O-groups \citep{DebRoy2016}.
There are potentially other subtype schemes that would show discrepancies between genetic data and subtype grouping.
A tool that can evaluate the ability of a genotype to predict a subtype would be better equipped for developing the new subtype schemes or updating current schemes for WGS workflows.
The \texttt{add} pipeline in Phylotyper tests subtype schemes for their predictive accuracy by returning an F-score based on a cross-validation assessment.
The validation verifies that the phylogeny generated by the training set sequences can be used to predict the subtypes with a high level of accuracy.
We showed in empirical tests that the farther a subtype is dissociated from the input gene's phylogeny, the lower the F-score computed in the Phylotyper \texttt{add} pipeline.
In addition to this subtype-level verification, Phylotyper also computes confidence scores for each individual prediction that reflect the rate of subtype change occurring at the input gene's phylogenetic locale.
If an input sequence falls in a region in the phylogenetic tree where the subtype is highly fluid due to evolution or poor subtype-genotype correlation, users would be notified in the confidence score and in the phylogenetic tree visualization output by Phylotyper. 
The ability of Phylotyper to inform users about the level of agreement between the subtype assignments and genotype makes it uniquely capable of transitioning historical subtype data to new whole genome sequence-based systems.\vspace*{-10pt}
